**Selective DNA and Protein Isolation from Marine Macrophyte Surfaces**

### 2020-11-19 submission/manuscript.Rmd
Typos corrected.

To evaluate whether the obtained DNA is suitable to determine the composition of the microbial community Illumina sequencing of the V4 region of the 16S rRNA was performed. Sequencing yielded a total of `r number_sequences` sequences after quality curation and exclusion of eukaryotic, mitochondrial and no relative sequences. The number of sequences classified as chloroplasts was `r number_chloroplast_sequences`. After excluding these sequences the total number of retrieved reads was `r number_sequences_without_chloroplast`, ranging from `r min_max$min` to `r min_max$max` sequences per sample (\autoref{supp-nseq_notus}). Even when the highest sequencing effort was applied, the rarefaction curves did not level off which is commonly observed in high-throughput 16S rRNA amplicon sequencing (\autoref{supp-rarefaction}). Sequences clustering at a similarity level of 97 \si{\percent} yielded a total of `r no_otus` different OTUs. Taxonomic classification of reads revealed a macrophyte-associated epiphytic community that mainly composed of *Alphaproteobacteria* (`r filter(plot, taxon=="Alphaproteobacteria")$mean` ± `r filter(plot, taxon=="Alphaproteobacteria")$sd` \si{\percent}), *Bacteroidota* (`r filter(plot, taxon=="Bacteroidota")$mean` ± `r filter(plot, taxon=="Bacteroidota")$sd` \si{\percent}), *Gammaproteobacteria* (`r filter(plot, taxon=="Gammaproteobacteria")$mean` ± `r filter(plot, taxon=="Gammaproteobacteria")$sd` \si{\percent}), *Desulfobacterota* (`r filter(plot, taxon=="Desulfobacterota")$mean` ± `r filter(plot, taxon=="Desulfobacterota")$sd` \si{\percent}), *Cyanobacteria* (`r filter(plot, taxon=="Cyanobacteria")$mean` ± `r filter(plot, taxon=="Cyanobacteria")$sd` \si{\percent}) and *Planctomycetota* (`r filter(plot, taxon=="Planctomycetota")$mean` ± `r filter(plot, taxon=="Planctomycetota")$sd` \si{\percent}) (\autoref{community}).

Primers used in this study, as in many other 16S rRNA amplicon procedures, also amplified chloroplast 16S rRNA genes. We observed a high proportion of chloroplast sequences in all analyzed samples (`r filter(plot, taxon=="Chloroplast")$mean` ± `r filter(plot, taxon=="Chloroplast")$sd` \si{\percent}) (\autoref{community}). To determine whether chloroplast sequences originate from the host or eukaryotic epiphytic organisms, we exported SILVA-classified chloroplast sequences and reclassified them using the RDP (Ribosomal Database Project) training set that allows for a more detailed chloroplast classification. The largest proportion of sequences was classified as Bacillariophyta (`r filter(chloroplast, taxon=="Bacillariophyta")$mean` ± `r filter(chloroplast, taxon=="Bacillariophyta")$sd` \si{\percent}) indicating that the DNA removal procedure resulted in only minor co-extracted quantities of host DNA (\autoref{chloroplast}). Chloroplast sequences classified as Streptophyta constituted `r streptophyta_16S$mean` ± `r streptophyta_16S$sd` \si{\percent} of all chloroplast sequences originating from *C. nodosa* samples, while sequences classified as Chlorophyta comprised only `r chlorophyta_16S$mean` ± `r chlorophyta_16S$sd` \si{\percent} of all chloroplast sequences associated with *C. cylindracea* samples.

To determine whether the extracted DNA can be used for metagenomic sequencing, four samples containing epiphytic DNA were selected and shotgun sequenced using an Illumina platform. Metagenomic sequencing yielded between `r format(min(metadata_metagenomic_statistics$raw_R1), big.mark=',')` and `r format(max(metadata_metagenomic_statistics$raw_R1), big.mark=',')` sequence pairs (\autoref{supp-metagenomic_statistics}). Obtained sequences were successfully assembled into contigs whose L50 ranged from `r format(min(metadata_metagenomic_statistics$L50), big.mark=',')` to `r format(max(metadata_metagenomic_statistics$L50), big.mark=',')` bp. In addition, predicted coding sequences were functionally annotated (`r format(min(metadata_metagenomic_statistics$eggnog), big.mark=',')` -- `r format(max(metadata_metagenomic_statistics$eggnog), big.mark=',')` annotated sequences; \autoref{cog}) and taxonomically classified. Functional annotation allowed for an assessment of the relative contribution of each COG (Clusters of Orthologous Groups) functional category to the total number of annotated coding sequences (\autoref{cog}a). Functional categories containing the highest number of sequences were `r cog_relative$"COG Functional Category"[3]` (`r cog_relative$Name[3]`), `r cog_relative$"COG Functional Category"[4]` (`r cog_relative$Name[4]`), `r cog_relative$"COG Functional Category"[5]` (`r cog_relative$Name[5]`), `r cog_relative$"COG Functional Category"[6]` (`r cog_relative$Name[6]`) and `r cog_relative$"COG Functional Category"[7]` (`r cog_relative$Name[7]`). If host DNA is co-extracted with epiphytes it should be detected in large proportions in sequenced metagenomes. However, no large proportions of coding sequences classified as Streptophyta and Chlorophyta were detected (\autoref{supp-metagenomic_taxonomy}). Sequenced metagenomic DNA originating from the surface of *C. nodosa* contained `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCyM" & phylum=="Streptophyta")$cds_relative` \si{\percent} of coding sequences classified as Streptophyta in December 2017 and `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCyM" & phylum=="Streptophyta")$cds_relative` \si{\percent} in June 2018. Furthermore, the summed RPKM (Reads Per Kilobase Million) of these sequences constituted `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCyM" & phylum=="Streptophyta")$rpkm_relative` \si{\percent} of total RPKM of all successfully classified sequences in December 2017 and `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCyM" & phylum=="Streptophyta")$rpkm_relative` \si{\percent} in June 2018. Similar low proportions of host coding sequences were detected in metagenomic samples originating from the surfaces of *C. cylindracea*. Of all successfully classified coding sequences `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCa" & phylum=="Chlorophyta")$cds_relative` \si{\percent} were classified as Chlorophyta in December 2017 and `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCa" & phylum=="Chlorophyta")$cds_relative` \si{\percent} in June 2018. A relatively higher proportion of RPKM of these sequences than in the case of *C. nodosa* was observed, indicating a higher coextraction of host DNA in *C. cylindracea*. In December, the proportion of RPKM of sequences classified as Chlorophyta was `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCa" & phylum=="Chlorophyta")$rpkm_relative` \si{\percent}, while in June 2018 it reached `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCa" & phylum=="Chlorophyta")$rpkm_relative` \si{\percent}.

@@

## Discussion
To test whether the developed DNA and protein isolation protocols efficiently detach microbes from the macrophyte surface we selected *C. nodosa* and *C. cylindracea* as representatives of seagrass and macroalgal species. These species differ morphologically. While *C. nodosa* leaves are flat, *C. cylindracea* thalli are characterized by an uneven surface [@Kuo2001; @Verlaque2003]. The developed protocol led to an almost complete removal of epiphytic cells from the surfaces of both species comparable to the result of @Burke2009, indicating that structural differences do not impact the removal efficiency. In addition, isolation protocols were tested in two contrasting seasons, as it is known that macrophytes are harboring more algal epiphytes during autumn and winter [@Reyes2001]. No differences in the removal efficiency was observed between seasons suggesting that these protocols can be used on macrophyte samples retrieved throughout the year. Also, no removal differences were observed on samples derived from the same host but from different locations.

@@

### DNA isolation
The DNA was isolated according to the protocol for isolation from filters described in @Massana1997. This protocol was modified and adapted for microbial DNA isolation from macrophyte surfaces as described below. Five (5) \si{\ml} of lysis buffer (40 \si{\milli\Molar} EDTA, 50 \si{\milli\Molar} Tris-HCl, 0.75 \si{\Molar} sucrose; pH 8.3) was added to 1 \si{\g} wet weight of leaves or 2 \si{\g} wet-weight of thalli. For every sample, duplicate isolations were performed. Lysozyme was added (final concentration 1 \si{\mg\per\ml}) and the mixture was incubated at 37 \si{\degreeCelsius} for 30 \si{\minute}. Subsequently, proteinase K (final concentration 0.5 \si{\mg\per\ml}) and SDS (final concentration 1 \si{\percent}) were added and the samples were incubated at 55 \si{\degreeCelsius} for 2 \si{\hour}. Following the incubation, tubes were vortexed for 10 \si{\minute} and the mixture containing lysed epiphytic cells was separated from host leaves or thalli by transferring the solution into a clean tube. The lysate was extracted twice with a mixture of phenol:chloroform:isoamyl alcohol (25:24:1; pH 8) and once with chloroform:isoamyl alcohol (24:1). After each addition of an organic solvent mixture, tubes were slightly vortexed and centrifuged at 4,500 × g for 10 \si{\minute}. Following each centrifugation the aqueous phases were retrieved. After the final extraction 1/10 of chilled 3 \si{\Molar} sodium acetate (pH 5.2) was added. DNA was precipitated by adding 1 volume of chilled isopropanol, incubating the mixtures overnight at \num{-20} \si{\degreeCelsius} and centrifuging at 16,000 × g and 4 \si{\degreeCelsius} for 20 \si{\minute}. The pellet was washed twice with 1 \si{\ml} of chilled 70 \si{\percent} ethanol and centrifuged after each washing step at 20,000 × g and 4 \si{\degreeCelsius} for 10 \si{\minute}. After the first washing step duplicate pellets from the same sample were pooled and transferred to a clean 1.5 \si{\ml} tube. The dried pellet was resuspended in 100 \si{\ul} of deionized water.

### Illumina 16S rRNA sequencing
An aliquot of isolated DNA was treated with RNase A (final concentration 200 \si{\ug\per\ml}) for 2 \si{\hour} at 37 \si{\degreeCelsius}. The DNA concentration was determined using the Quant-iT PicoGreen dsDNA Assay Kit (Invitrogen, USA) according to the manufacturer’s instructions and diluted to 1 \si{\ng\per\ul}. The V4 region of the 16S rRNA gene was amplified using a two-step PCR procedure. In the first PCR the 515F (5'-GTGYCAGCMGCCGCGGTAA-3') and 806R (5'-GGACTACNVGGGTWTCTAAT-3') primers from the Earth Microbiome Project (http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/) were used to amplify the target region [@Caporaso2012; @Apprill2015; @Parada2016]. These primers contained on their 5' end a tagged sequence. Each sample was amplified in four parallel 25 \si{\ul} reactions of which each contained 1 × Q5 Reaction Buffer, 0.2 \si{\milli\Molar} of dNTPmix, 0.7 \si{\mg\per\ml} BSA (Bovine Serum Albumin), 0.2 \si{\micro\Molar} of forward and reverse primers, 0.5 U of Q5 High-Fidelity DNA Polymerase (New England Biolabs, USA) and 5 \si{\ng} of DNA template. Cycling conditions were: initial denaturation at 94 \si{\degreeCelsius} for 3 \si{\minute}, 20 cycles of denaturation at 94 \si{\degreeCelsius} for 45 \si{\s}, annealing at 50 \si{\degreeCelsius} for 60 \si{\s} and elongation at 72 \si{\degreeCelsius} for 90 \si{\s}, finalized by an elongation step at 72 \si{\degreeCelsius} for 10 \si{\minute}. The four parallel reactions volumes were pooled and PCR products were purified using the GeneJET PCR Purification Kit (Thermo Fisher Scientific, USA) according to the manufacturer's instructions and following the protocol that included isopropanol addition for better small DNA fragment yield. The column was eluted in 30 \si{\ul} of deionized water. Purified PCR products were sent for Illumina MiSeq sequencing (2 × 250 bp) at IMGM Laboratories, Martinsried, Germany. Before sequencing at IMGM, the second PCR amplification of the two-step PCR procedure was performed using primers targeting the tagged region incorporated in the first PCR. In addition, these primers contained adapter and sample-specific index sequences. The second PCR was carried out for 8 cycles. Beside samples, a positive and negative control were sequenced. A negative control was comprised of four parallel PCR reactions without DNA template, while for a positive control a mock community composed of evenly mixed DNA material originating from 20 bacterial strains (ATCC MSA-1002, ATCC, USA) was used. Partial 16S rRNA sequences obtained in this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession numbers SAMEA6786270, SAMEA6648792 -- SAMEA6648794, SAMEA6648809 -- SAMEA6648811.

@@

Obtained sequences were analyzed on the computer cluster Isabella (University Computing Center, University of Zagreb) using mothur (version 1.43.0) [@Schloss2009] according to the MiSeq Standard Operating Procedure (MiSeq SOP; https://mothur.org/wiki/MiSeq_SOP) [@Kozich2013] and recommendations given from the Riffomonas project to enhance data reproducibility (http://www.riffomonas.org/). For alignment and classification of sequences the SILVA SSU Ref NR 99 database (release 138; http://www.arb-silva.de) was used [@Quast2013; @Yilmaz2014]. Sequences classified as chloroplasts by SILVA were exported and reclassified using mothur and the RDP (http://rdp.cme.msu.edu/) training set (version 16) reference files adapted for mothur [@Cole2014]. In comparison to SILVA, RDP allows a more detailed classification of chloroplast sequences. Based on the ATCC MSA-1002 mock community included in the analysis a sequencing error rate of `r error` \si{\percent} was determined, which is in line with previously reported values for next-generation sequencing data [@Kozich2013; @Schloss2016]. In addition, the negative control processed together with the samples yielded only `r nc` sequences after sequence quality curation.

@@

Obtained sequences were analyzed on the Life Science Compute Cluster (LiSC) (CUBE -- Computational Systems Biology, University of Vienna). Individual sequences were assembled using MEGAHIT (version  1.1.2) [@Li2015] under default settings. Putative genes were predicted from contings longer than 200 bp using Prodigal (version  2.6.3) [@Hyatt2010] in metagenome mode (-p meta). Abundances of predicted genes were expressed as Reads Per Kilobase Million (RPKM) and calculated using the BWA algorithm (version  0.7.16a) [@Li2009a]. All predicted genes were functionally annotated using the eggNOG-mapper [@Huerta-Cepas2017] and eggNOG database (version 5.0) [@Huerta-Cepas2019]. Sequence taxonomy classification was determined using the lowest common ancestor algorithm adapted from DIAMOND (version 0.8.36) [@Buchfink2015] and by searching against the NCBI non-redundant database (NR). To determine the phylogeny, the top 10 \si{\percent} hits with an e-value < 1 × 10^-5^ were used (\-\-top 10). Sequence renaming, coverage information computing and metagenomic statistics calculations were performed using software tools from BBTools (https://jgi.doe.gov/data-and-tools/bbtools). Metagenomic sequences obtained in this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession numbers SAMEA6648795, SAMEA6648797, SAMEA6648809 and SAMEA6648811.

@@

### Metaproteomics
Isolated proteins were whole trypsin digested using the FASP (filter-aided sample preparation) Protein Digestion Kit (Expedeon, UK) according to the manufacturer's instructions with small modifications [@Wisniewski2009]. Prior to loading the solution onto the column, protein pellets were solubilized in a urea sample buffer included in the kit amended with DTT (final concentration 100 \si{\milli\Molar}) for 45 \si{\minute} at room temperature and centrifuged at 20,000 × g for 2 -- 5 \si{\minute} at room temperature to remove larger particles. The first washing step after protein solution loading was repeated twice. In addition, the centrifugation steps were prolonged if the column was clogged. Trypsin digestion was performed on column filters at 37 \si{\degreeCelsius} overnight for 18 \si{\hour}. The final filtrate containing peptides was acidified with 1 \si{\percent} (final concentration) trifluoroacetic acid (TFA), freezed at \num{-80} \si{\degreeCelsius}, lyophilized and sent to VIME -- Vienna Metabolomics Center (University of Vienna) for metaproteomic analysis. Peptides were resuspended in 1 \si{\percent} (final concentration) TFA, desalted using the Pierce C18 Tips (Thermo Fisher Scientific, USA) according to the manufacturer's instructions and sequenced on a Q Exactive Hybrid Quadrupole-Orbitrap Mass Spectrometer (Thermo Fisher Scientific, USA). Obtained MS/MS spectra were searched against a protein database composed of combined sequenced metagenomes using SEQUEST-HT engines and validated with Percolator in Proteome Discoverer 2.1 (Thermo Fisher Scientific, USA). The target-decoy approach was used to reduce the probability of false peptide identification. Results whose false discovery rate at the peptide level was \SI{< 1}{\percent} were kept. For protein identification a minimum of two peptides and one unique peptide were required. For protein quantification, a chromatographic peak area-based free quantitative method was applied.

@@

### Confocal microscopy
Host leaves and thalli from DNA and protein isolation steps were washed seven times in deionized water and fixed with formaldehyde (final concentration ~ 3 \si{\percent}). In addition, nontreated leaves and thalli, washed three times in ASW to remove loosely attached microbial cells, were fixed in the same concentration of formaldehyde and used as a positive control. For-long term storage, fixed leaves and thalli were immersed in a mixture of phosphate-buffered saline (PBS) and ethanol (1:1) and stored at \num{-20} \si{\degreeCelsius}. Treated and untreated segments of leaves and thalli were stained in a 2 × solution of SYBR Green I and examined under a Leica TCS SP8 X FLIM confocal microscope (Leica Microsystems, Germany).

>>>>

To evaluate whether the obtained DNA is suitable to determine the composition of the microbial community Illumina sequencing of the V4 region of the 16S rRNA was performed. Sequencing yielded a total of `r number_sequences` sequences after quality curation and exclusion of eukaryotic, mitochondrial and no relative sequences. The number of sequences classified as chloroplasts was `r number_chloroplast_sequences`. After excluding these sequences, the total number of retrieved reads was `r number_sequences_without_chloroplast`, ranging from `r min_max$min` to `r min_max$max` sequences per sample (\autoref{supp-nseq_notus}). Even when the highest sequencing effort was applied, the rarefaction curves did not level off which is commonly observed in high-throughput 16S rRNA amplicon sequencing (\autoref{supp-rarefaction}). Sequences clustering at a similarity level of 97 \si{\percent} yielded a total of `r no_otus` different OTUs. Taxonomic classification of reads revealed a macrophyte-associated epiphytic community that mainly composed of *Alphaproteobacteria* (`r filter(plot, taxon=="Alphaproteobacteria")$mean` ± `r filter(plot, taxon=="Alphaproteobacteria")$sd` \si{\percent}), *Bacteroidota* (`r filter(plot, taxon=="Bacteroidota")$mean` ± `r filter(plot, taxon=="Bacteroidota")$sd` \si{\percent}), *Gammaproteobacteria* (`r filter(plot, taxon=="Gammaproteobacteria")$mean` ± `r filter(plot, taxon=="Gammaproteobacteria")$sd` \si{\percent}), *Desulfobacterota* (`r filter(plot, taxon=="Desulfobacterota")$mean` ± `r filter(plot, taxon=="Desulfobacterota")$sd` \si{\percent}), *Cyanobacteria* (`r filter(plot, taxon=="Cyanobacteria")$mean` ± `r filter(plot, taxon=="Cyanobacteria")$sd` \si{\percent}) and *Planctomycetota* (`r filter(plot, taxon=="Planctomycetota")$mean` ± `r filter(plot, taxon=="Planctomycetota")$sd` \si{\percent}) (\autoref{community}).

Primers used in this study, as in many other 16S rRNA amplicon procedures, also amplified chloroplast 16S rRNA genes. We observed a high proportion of chloroplast sequences in all analysed samples (`r filter(plot, taxon=="Chloroplast")$mean` ± `r filter(plot, taxon=="Chloroplast")$sd` \si{\percent}) (\autoref{community}). To determine whether chloroplast sequences originate from the host or eukaryotic epiphytic organisms, we exported SILVA-classified chloroplast sequences and reclassified them using the RDP (Ribosomal Database Project) training set that allows for a more detailed chloroplast classification. The largest proportion of sequences was classified as Bacillariophyta (`r filter(chloroplast, taxon=="Bacillariophyta")$mean` ± `r filter(chloroplast, taxon=="Bacillariophyta")$sd` \si{\percent}) indicating that the DNA removal procedure resulted in only minor co-extracted quantities of host DNA (\autoref{chloroplast}). Chloroplast sequences classified as Streptophyta constituted `r streptophyta_16S$mean` ± `r streptophyta_16S$sd` \si{\percent} of all chloroplast sequences originating from *C. nodosa* samples, while sequences classified as Chlorophyta comprised only `r chlorophyta_16S$mean` ± `r chlorophyta_16S$sd` \si{\percent} of all chloroplast sequences associated with *C. cylindracea* samples.

To determine whether the extracted DNA can be used for metagenomic sequencing, four samples containing epiphytic DNA were selected and shotgun sequenced using an Illumina platform. Metagenomic sequencing yielded between `r format(min(metadata_metagenomic_statistics$raw_R1), big.mark=',')` and `r format(max(metadata_metagenomic_statistics$raw_R1), big.mark=',')` sequence pairs (\autoref{supp-metagenomic_statistics}). Obtained sequences were successfully assembled into contigs whose L50 ranged from `r format(min(metadata_metagenomic_statistics$L50), big.mark=',')` to `r format(max(metadata_metagenomic_statistics$L50), big.mark=',')` bp. In addition, predicted coding sequences were functionally annotated (`r format(min(metadata_metagenomic_statistics$eggnog), big.mark=',')` -- `r format(max(metadata_metagenomic_statistics$eggnog), big.mark=',')` annotated sequences; \autoref{cog}a) and taxonomically classified. Functional annotation allowed for an assessment of the relative contribution of each COG (Clusters of Orthologous Groups) functional category to the total number of annotated coding sequences (\autoref{cog}a). Functional categories containing the highest number of sequences were `r cog_relative$"COG Functional Category"[3]` (`r cog_relative$Name[3]`), `r cog_relative$"COG Functional Category"[4]` (`r cog_relative$Name[4]`), `r cog_relative$"COG Functional Category"[5]` (`r cog_relative$Name[5]`), `r cog_relative$"COG Functional Category"[6]` (`r cog_relative$Name[6]`) and `r cog_relative$"COG Functional Category"[7]` (`r cog_relative$Name[7]`). If host DNA is co-extracted with epiphytes it should be detected in large proportions in sequenced metagenomes. However, no large proportions of coding sequences classified as Streptophyta and Chlorophyta were detected (\autoref{supp-metagenomic_taxonomy}). Sequenced metagenomic DNA originating from the surface of *C. nodosa* contained `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCyM" & phylum=="Streptophyta")$cds_relative` \si{\percent} of coding sequences classified as Streptophyta in December 2017 and `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCyM" & phylum=="Streptophyta")$cds_relative` \si{\percent} in June 2018. Furthermore, the summed RPKM (Reads Per Kilobase Million) of these sequences constituted `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCyM" & phylum=="Streptophyta")$rpkm_relative` \si{\percent} of total RPKM of all successfully classified sequences in December 2017 and `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCyM" & phylum=="Streptophyta")$rpkm_relative` \si{\percent} in June 2018. Similar low proportions of host coding sequences were detected in metagenomic samples originating from the surfaces of *C. cylindracea*. Of all successfully classified coding sequences `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCa" & phylum=="Chlorophyta")$cds_relative` \si{\percent} were classified as Chlorophyta in December 2017 and `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCa" & phylum=="Chlorophyta")$cds_relative` \si{\percent} in June 2018. A relatively higher proportion of RPKM of these sequences than in the case of *C. nodosa* was observed, indicating a higher co-extraction of host DNA in *C. cylindracea*. In December, the proportion of RPKM of sequences classified as Chlorophyta was `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCa" & phylum=="Chlorophyta")$rpkm_relative` \si{\percent}, while in June 2018 it reached `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCa" & phylum=="Chlorophyta")$rpkm_relative` \si{\percent}.

@@

## Discussion
To test whether the developed DNA and protein isolation protocols efficiently detach microbes from the macrophyte surface we selected *C. nodosa* and *C. cylindracea* as representatives of seagrass and macroalgal species. These species differ morphologically. While *C. nodosa* leaves are flat, *C. cylindracea* thalli are characterized by an uneven surface [@Kuo2001; @Verlaque2003]. The developed protocol led to an almost complete removal of epiphytic cells from the surfaces of both species comparable to the result of @Burke2009, indicating that structural differences do not impact the removal efficiency. In addition, isolation protocols were tested in two contrasting seasons, as it is known that macrophytes are harbouring more algal epiphytes during autumn and winter [@Reyes2001]. No differences in the removal efficiency was observed between seasons suggesting that these protocols can be used on macrophyte samples retrieved throughout the year. Also, no removal differences were observed on samples derived from the same host but from different locations.

@@

### DNA isolation
The DNA was isolated according to the protocol for isolation from filters described in @Massana1997. This protocol was modified and adapted for microbial DNA isolation from macrophyte surfaces as described below. Five (5) \si{\ml} of lysis buffer (40 \si{\milli\Molar} EDTA, 50 \si{\milli\Molar} Tris-HCl, 0.75 \si{\Molar} sucrose; pH 8.3) was added to 1 \si{\g} wet weight of leaves or 2 \si{\g} wet-weight of thalli. For every sample, duplicate isolations were performed. Lysozyme was added (final concentration 1 \si{\mg\per\ml}) and the mixture was incubated at 37 \si{\degreeCelsius} for 30 \si{\minute}. Subsequently, proteinase K (final concentration 0.5 \si{\mg\per\ml}) and SDS (final concentration 1 \si{\percent}) were added and the samples were incubated at 55 \si{\degreeCelsius} for 2 \si{\hour}. Following the incubation, tubes were vortexed for 10 \si{\minute} and the mixture containing lysed epiphytic cells was separated from host leaves or thalli by transferring the solution into a clean tube. The lysate was extracted twice with a mixture of phenol:chloroform:isoamyl alcohol (25:24:1; pH 8) and once with chloroform:isoamyl alcohol (24:1). After each addition of an organic solvent mixture, tubes were slightly vortexed and centrifuged at 4,500 × g for 10 \si{\minute}. Following each centrifugation the aqueous phases were retrieved. After the final extraction 1/10 of chilled 3 \si{\Molar} sodium acetate (pH 5.2) was added. DNA was precipitated by adding 1 volume of chilled isopropanol, incubating the mixtures overnight at \num{-20} \si{\degreeCelsius} and centrifuging at 16,000 × g and 4 \si{\degreeCelsius} for 20 \si{\minute}. The pellet was washed twice with 1 \si{\ml} of chilled 70 \si{\percent} ethanol and centrifuged after each washing step at 20,000 × g and 4 \si{\degreeCelsius} for 10 \si{\minute}. After the first washing step duplicate pellets from the same sample were pooled and transferred to a clean 1.5 \si{\ml} tube. The dried pellet was re-suspended in 100 \si{\ul} of deionized water.

### Illumina 16S rRNA sequencing
An aliquot of isolated DNA was treated with RNase A (final concentration 200 \si{\ug\per\ml}) for 2 \si{\hour} at 37 \si{\degreeCelsius}. The DNA concentration was determined using the Quant-iT PicoGreen dsDNA Assay Kit (Invitrogen, USA) according to the manufacturer’s instructions and diluted to 1 \si{\ng\per\ul}. The V4 region of the 16S rRNA gene was amplified using a two-step PCR procedure. In the first PCR the 515F (5\prime-GTGYCAGCMGCCGCGGTAA-3\prime) and 806R (5\prime-GGACTACNVGGGTWTCTAAT-3\prime) primers from the Earth Microbiome Project (http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/) were used to amplify the target region [@Caporaso2012; @Apprill2015; @Parada2016]. These primers contained on their 5\prime end a tagged sequence. Each sample was amplified in four parallel 25 \si{\ul} reactions of which each contained 1 × Q5 Reaction Buffer, 0.2 \si{\milli\Molar} of dNTPmix, 0.7 \si{\mg\per\ml} BSA (Bovine Serum Albumin), 0.2 \si{\micro\Molar} of forward and reverse primers, 0.5 U of Q5 High-Fidelity DNA Polymerase (New England Biolabs, USA) and 5 \si{\ng} of DNA template. Cycling conditions were: initial denaturation at 94 \si{\degreeCelsius} for 3 \si{\minute}, 20 cycles of denaturation at 94 \si{\degreeCelsius} for 45 \si{\s}, annealing at 50 \si{\degreeCelsius} for 60 \si{\s} and elongation at 72 \si{\degreeCelsius} for 90 \si{\s}, finalized by an elongation step at 72 \si{\degreeCelsius} for 10 \si{\minute}. The four parallel reactions volumes were pooled and PCR products were purified using the GeneJET PCR Purification Kit (Thermo Fisher Scientific, USA) according to the manufacturer's instructions and following the protocol that included isopropanol addition for better small DNA fragment yield. The column was eluted in 30 \si{\ul} of deionized water. Purified PCR products were sent for Illumina MiSeq sequencing (2 × 250 bp) at IMGM Laboratories, Martinsried, Germany. Before sequencing at IMGM, the second PCR amplification of the two-step PCR procedure was performed using primers targeting the tagged region incorporated in the first PCR. In addition, these primers contained adapter and sample-specific index sequences. The second PCR was carried out for 8 cycles. Beside samples, a positive and negative control were sequenced. A negative control was comprised of four parallel PCR reactions without DNA template, while for a positive control a mock community composed of evenly mixed DNA material originating from 20 bacterial strains (ATCC MSA-1002, ATCC, USA) was used. Partial 16S rRNA sequences obtained in this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession numbers SAMEA6786270, SAMEA6648792 -- SAMEA6648794, SAMEA6648809 -- SAMEA6648811.

@@

Obtained sequences were analysed on the computer cluster Isabella (University Computing Center, University of Zagreb) using mothur (version 1.43.0) [@Schloss2009] according to the MiSeq Standard Operating Procedure (MiSeq SOP; https://mothur.org/wiki/MiSeq_SOP) [@Kozich2013] and recommendations given from the Riffomonas project to enhance data reproducibility (http://www.riffomonas.org/). For alignment and classification of sequences the SILVA SSU Ref NR 99 database (release 138; http://www.arb-silva.de) was used [@Quast2013; @Yilmaz2014]. Sequences classified as chloroplasts by SILVA were exported and reclassified using mothur and the RDP (http://rdp.cme.msu.edu/) training set (version 16) reference files adapted for mothur [@Cole2014]. In comparison to SILVA, RDP allows a more detailed classification of chloroplast sequences. Based on the ATCC MSA-1002 mock community included in the analysis a sequencing error rate of `r error` \si{\percent} was determined, which is in line with previously reported values for next-generation sequencing data [@Kozich2013; @Schloss2016]. In addition, the negative control processed together with the samples yielded only `r nc` sequences after sequence quality curation.

@@

Obtained sequences were analysed on the Life Science Compute Cluster (LiSC) (CUBE -- Computational Systems Biology, University of Vienna). Individual sequences were assembled using MEGAHIT (version  1.1.2) [@Li2015] under default settings. Putative genes were predicted from contigs longer than 200 bp using Prodigal (version  2.6.3) [@Hyatt2010] in metagenome mode (-p meta). Abundances of predicted genes were expressed as Reads Per Kilobase Million (RPKM) and calculated using the BWA algorithm (version  0.7.16a) [@Li2009a]. All predicted genes were functionally annotated using the eggNOG-mapper [@Huerta-Cepas2017] and eggNOG database (version 5.0) [@Huerta-Cepas2019]. Sequence taxonomy classification was determined using the lowest common ancestor algorithm adapted from DIAMOND (version 0.8.36) [@Buchfink2015] and by searching against the NCBI non-redundant database (NR). To determine the phylogeny, the top 10 \si{\percent} hits with an e-value < 1 × 10^-5^ were used (\-\-top 10). Sequence renaming, coverage information computing and metagenomic statistics calculations were performed using software tools from BBTools (https://jgi.doe.gov/data-and-tools/bbtools). Metagenomic sequences obtained in this study have been deposited in the European Nucleotide Archive (ENA) at EMBL-EBI under accession numbers SAMEA6648795, SAMEA6648797, SAMEA6648809 and SAMEA6648811.

@@

### Metaproteomics
Isolated proteins were whole trypsin digested using the FASP (filter-aided sample preparation) Protein Digestion Kit (Expedeon, UK) according to the manufacturer's instructions with small modifications [@Wisniewski2009]. Prior to loading the solution onto the column, protein pellets were solubilized in a urea sample buffer included in the kit amended with DTT (final concentration 100 \si{\milli\Molar}) for 45 \si{\minute} at room temperature and centrifuged at 20,000 × g for 2 -- 5 \si{\minute} at room temperature to remove larger particles. The first washing step after protein solution loading was repeated twice. In addition, the centrifugation steps were prolonged if the column was clogged. Trypsin digestion was performed on column filters at 37 \si{\degreeCelsius} overnight for 18 \si{\hour}. The final filtrate containing peptides was acidified with 1 \si{\percent} (final concentration) trifluoroacetic acid (TFA), freezed at \num{-80} \si{\degreeCelsius}, lyophilized and sent to VIME -- Vienna Metabolomics Center (University of Vienna) for metaproteomic analysis. Peptides were re-suspended in 1 \si{\percent} (final concentration) TFA, desalted using the Pierce C18 Tips (Thermo Fisher Scientific, USA) according to the manufacturer's instructions and sequenced on a Q Exactive Hybrid Quadrupole-Orbitrap Mass Spectrometer (Thermo Fisher Scientific, USA). Obtained MS/MS spectra were searched against a protein database composed of combined sequenced metagenomes using SEQUEST-HT engines and validated with Percolator in Proteome Discoverer 2.1 (Thermo Fisher Scientific, USA). The target-decoy approach was used to reduce the probability of false peptide identification. Results whose false discovery rate at the peptide level was \SI{< 1}{\percent} were kept. For protein identification a minimum of two peptides and one unique peptide were required. For protein quantification, a chromatographic peak area-based free quantitative method was applied.

@@

### Confocal microscopy
Host leaves and thalli from DNA and protein isolation steps were washed seven times in deionized water and fixed with formaldehyde (final concentration ~3 \si{\percent}). In addition, non-treated leaves and thalli, washed three times in ASW to remove loosely attached microbial cells, were fixed in the same concentration of formaldehyde and used as a positive control. For-long term storage, fixed leaves and thalli were immersed in a mixture of phosphate-buffered saline (PBS) and ethanol (1:1) and stored at \num{-20} \si{\degreeCelsius}. Treated and untreated segments of leaves and thalli were stained in a 2 × solution of SYBR Green I and examined under a Leica TCS SP8 X FLIM confocal microscope (Leica Microsystems, Germany).
